---
title: "Light in the water column"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Light in the water column}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4
)
```

A researcher has deployed irradiance sensors at three depths on a coastal reef and
wants to quantify how the light field changes with depth — both from direct
measurements and extended to greater depths using a standard optical water-type model.

## Loading field data

The `Naples` dataset ships with luxR: spectrally resolved photon-flux irradiance
measured at 0, 5, and 10 m depth at a coastal Mediterranean site. `from_naples()`
wraps it in a `lux_spectrum` object that carries the wavelengths, unit, and metadata.

```{r load}
library(luxR)

x0m  <- from_naples("0m")
x5m  <- from_naples("5m")
x10m <- from_naples("10m")

x0m
```

```{r plot-surface}
plot(x0m, main = "Surface irradiance — Naples coastal site")
```

The spectrum peaks in the blue-green window typical of coastal Mediterranean water.

> **Numeric API note:** All functions in luxR also accept plain numeric vectors.
> For example, `par_irradiance(Naples$depth_0m, Naples$wv, photonic = TRUE)`.
> See each function's help page for the numeric signature.

## Converting units

The Naples data are stored in µmol photons m⁻² s⁻¹ nm⁻¹. For Beer-Lambert
propagation and PAR integration we work in energy units; `convert_unit()` handles
the conversion.

```{r convert}
x0m_W  <- convert_unit(x0m,  "W/m2/nm")
x5m_W  <- convert_unit(x5m,  "W/m2/nm")
x10m_W <- convert_unit(x10m, "W/m2/nm")

x0m_W
```

### Validation against photobiology

luxR's energy ↔ photon conversion is the elementary radiometric identity
$n(\lambda) = E(\lambda)\,\lambda / (h c)$. The same constant underlies the
[photobiology](https://CRAN.R-project.org/package=photobiology) package's
`e2quantum_multipliers()`, so the two agree to machine precision — a check on
luxR's radiometric constants that runs automatically wherever photobiology is
installed (`tests/testthat/test-photobiology-validation.R`).

The same guarded test file independently integrates flat and smooth blue,
green, and red spectra with `photobiology::response()` and compares them with
`irradiance2lux()` and `scotopic_lux()`. Regular 1 nm and irregular wavelength
grids, plus the bundled Naples field spectrum, are covered. Both packages
receive the same checksum-pinned official CIE 018:2019 response tables in these
tests, so the benchmark isolates numerical weighting and integration rather
than comparing different observer standards.

```{r photobiology-validation, eval = requireNamespace("photobiology", quietly = TRUE)}
lam <- seq(300, 800, by = 50)
pb  <- photobiology::e2quantum_multipliers(lam, molar = FALSE)
tab <- data.frame(
  wavelength_nm = lam,
  ratio         = W2photon(1, lam) / pb,
  rel_diff      = abs(W2photon(1, lam) - pb) / pb
)
knitr::kable(tab, digits = c(0, 12, 16),
             col.names = c("λ (nm)", "luxR / photobiology", "rel. difference"))
```

## PAR across depth

Photosynthetically active radiation (PAR) is the photon flux integrated over
400–700 nm, expressed in µmol m⁻² s⁻¹. `par_irradiance()` computes it directly
from an energy-based `lux_spectrum`.

```{r par}
par_table <- data.frame(
  depth_m  = c(0, 5, 10),
  PAR_umol = c(par_irradiance(x0m_W),
               par_irradiance(x5m_W),
               par_irradiance(x10m_W))
)
knitr::kable(par_table, digits = 1,
             col.names = c("Depth (m)", "PAR (µmol m⁻² s⁻¹)"))
```

## Fitting attenuation from measurements

`fit_Kd()` estimates the diffuse attenuation coefficient Kd(λ) from two depth
measurements. Kd is wavelength-dependent: low in the blue window and high in the red.

```{r fit-kd}
Kd_meas <- fit_Kd(x0m_W, 0, x10m_W, 10)

plot(x0m_W$lambda, Kd_meas, type = "l",
     xlab = "Wavelength (nm)",
     ylab = expression(K[d] ~ (m^{-1})),
     main = "Diffuse attenuation spectrum (0 and 10 m measurements)")
```

Kd is highest in the short-wavelength (violet) and red, and lowest near 490 nm — the wavelength
that penetrates deepest in coastal water.

## Extending to depth with a Jerlov model

Measurements reach only 10 m. To go deeper we use a Jerlov Type II coastal
water-type model and propagate the surface spectrum.

This is luxR's principal lightweight modeling tier. At each wavelength it uses
$E(z,\lambda)=E(z_0,\lambda)\exp[-K_d(\lambda)(z-z_0)]$. The calculation
assumes that $K_d(\lambda)$ is constant over the whole path and that wavelength
bins propagate independently. It is not an angular or multiple-scattering
radiative-transfer solution and does not represent fluorescence, Raman
scattering, or other wavelength redistribution. The bundled Jerlov table is
supported from 350–700 nm, which is why the input is explicitly restricted
below. There is no universal maximum valid depth: the selected water type must
remain representative of the modeled water column.

```{r jerlov}
x0m_Jerlov <- x0m_W[350, 700]
lam   <- x0m_Jerlov$lambda
Kd_II <- jerlov_Kd("II", lam)

specs <- propagate_spectrum(x0m_Jerlov, Kd_II, from = 0,
                            to = c(0, 5, 10, 20, 50))
plot(specs, main = "Depth-propagated spectra — Jerlov Type II")
```

Spectral narrowing with depth is clearly visible: red wavelengths are attenuated
first, leaving blue-green light dominant below 20 m.

Propagating a measured deep spectrum upward uses the same homogeneous-column
assumption in reverse. That reconstruction can strongly amplify measurement
noise and should not be interpreted as a direct surface observation.

## Band-integrated irradiance

The blue channel (400–500 nm) drives many biological processes including circadian
entrainment and short-wavelength photoreception. `band_irradiance()` integrates the
propagated spectrum over this window at each target depth.

```{r band}
bi <- band_irradiance(x0m_Jerlov, Kd_II,
                      depths     = c(0, 5, 10, 20, 50),
                      lambda_min = 400, lambda_max = 500)
knitr::kable(bi, digits = 3,
             col.names = c("Depth (m)", "Blue-band irradiance (W m⁻²)"))
```

## Photic depth

The photic zone — where photosynthesis is feasible — is conventionally defined as
the depth at which irradiance falls to 1% of the surface value. Using the
spectrally averaged Kd within the PAR window:

```{r photic}
Kd_PAR_mean <- mean(Kd_II[lam >= 400 & lam <= 700])
photic_z    <- photic_depth(Kd_PAR_mean)
cat(sprintf("1%% light level (Jerlov Type II): %.1f m\n", photic_z))
```
