---
title: "Polarization of the underwater light field"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polarization of the underwater light field}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4
)
library(luxR)
```

Sunlight scattered by water becomes partially **linearly polarized**, and many
aquatic animals — cephalopods, stomatopods, many crustaceans and some fish — see
that polarization and use it for navigation, contrast enhancement, and
communication. luxR provides a tunable, Rayleigh-like single-scattering
approximation to the underwater polarization field: enough to build intuition
and explore geometry, but **not** a vector radiative-transfer solution.

> **Model status:** `underwater_polarization()` is an uncalibrated approximation,
> not a quantitative prediction. Its result carries a `luxR.polarization`
> attribute recording that status and the exact parameter values. Quantitative
> use requires calibration for the relevant wavelength, water type, surface
> state, depth, and viewing geometry.

## The model

For singly (Rayleigh-like) scattered light, the **degree of polarization** (DoP)
depends on the scattering angle $\Theta$ between the line of sight and the
refracted solar beam,

$$ DoP(\Theta) = DoP_{max}\,\frac{\sin^2\Theta}{1 + \cos^2\Theta}, $$

peaking at $\Theta = 90^\circ$ and vanishing toward the sun and anti-sun. With
depth, multiple scattering randomizes the field, so DoP relaxes toward an
asymptotic floor. The sun's beam is refracted at the surface into Snell's window
first:

```{r refraction}
refracted_solar_angle(c(0, 30, 60, 90))   # in-water solar zenith (deg)
```

## Degree of polarization across the sky

With the sun overhead, polarization is strongest when looking horizontally
(90° from the sun) and disappears looking straight up at the sun:

```{r dop-zenith}
vz  <- seq(0, 90, by = 2)
res <- underwater_polarization(view_zenith = vz, sun_zenith = 0, depth = 1)

plot(res$scattering_angle, res$dop, type = "l", lwd = 2, col = "steelblue4",
     xlab = "Scattering angle from sun (deg)",
     ylab = "Degree of polarization",
     main = "Underwater DoP vs angle from the sun (near surface)")
abline(v = 90, lty = 3, col = "grey50")
```

## How polarization fades with depth

Looking at the 90° band (maximally polarized at the surface), DoP relaxes toward
the deep asymptotic value as multiple scattering takes over:

```{r dop-depth}
depths <- seq(0, 100, by = 2)
dz <- underwater_polarization(view_zenith = 90, sun_zenith = 0,
                              depth = depths, dop_max = 0.4, dop_asym = 0.03)

plot(depths, dz$dop, type = "l", lwd = 2, col = "firebrick",
     xlab = "Depth (m)", ylab = "Degree of polarization",
     main = "DoP at the 90-degree band vs depth")
abline(h = 0.03, lty = 3, col = "grey50")
```

The `dop_max`, `dop_asym`, and `z_p` arguments let you tune the approximation to
a water type or to field measurements. Their defaults are illustrative:

| Parameter | Default | Evidence and valid domain |
|---|---:|---|
| `dop_max` | 0.4 | Upper end of the roughly 0.25–0.40 maxima reported by Cronin & Shashar (2001) at 15 m in clear tropical marine water. The function defines this parameter at depth zero, so the observation is context rather than a calibration. |
| `dop_asym` | 0.03 | Heuristic deep-field floor. It has no empirical universal domain and must be calibrated for quantitative use. |
| `z_p` | 20 m | Heuristic relaxation scale, not a measured attenuation length. It must be calibrated for quantitative use. |
| `n` | 1.333 | Conventional visible-light approximation for water. Refractive index varies with wavelength, temperature, and salinity. |

Cronin & Shashar measured clear tropical reef water at 15 m over 350–600 nm and
found the strongest polarization 60–90° from the sun. The package tests this
angular pattern as a literature-consistency benchmark. It does not treat that
qualitative agreement as calibration.

## The e-vector

The angle of polarization (`aop`) is perpendicular to the scattering plane. It is
returned alongside DoP and is `NA` when looking straight along the solar beam,
where the e-vector is undefined:

```{r aop}
underwater_polarization(view_zenith = c(0, 45, 90), view_azimuth = 0,
                        sun_zenith = 30, sun_azimuth = 0, depth = 5)
```

For Stokes-vector inputs you can also compute the descriptors directly with
`degree_of_polarization()` and `angle_of_polarization()`. Invalid Stokes states
whose polarized intensity exceeds total intensity are rejected rather than
silently returning a degree greater than one.
