---
title: "Seeing through water: from sunlight to detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seeing through water: from sunlight to detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4
)
```

This vignette runs a single question end to end: **can a fish, at depth, detect a
reddish object against the open-water background — and how far away?** It chains
the three stages luxR is built around:

1. **Sunlight to depth** — cross the air–water surface and attenuate the solar
   spectrum through a Jerlov water type.
2. **Light to capture** — what the fish's photoreceptors actually catch, allowing
   for self-screening.
3. **Capture to detection** — the inherent object/background contrast and the
   distance at which it falls below threshold.

```{r load}
library(luxR)
```

## 1. Sunlight to depth

Start with the clear-noon solar spectrum just above the surface. A flat air–water
interface reflects a few percent of it away (Fresnel); `surface_transmittance()`
gives the fraction that actually enters the water.

```{r surface}
sp    <- solar_irradiance("clear_noon")   # W/m2/nm, above surface
lam   <- sp$wavelength

tau   <- surface_transmittance(angle = 30)            # sun 30 deg from vertical
E_sub <- sp$irradiance * tau                          # just below the surface
round(tau, 3)
```

Now propagate that subsurface spectrum down through Jerlov type II water with
Beer–Lambert attenuation. The water is a strong spectral filter: red light is
gone within a few metres, leaving a blue-green field at depth.

```{r propagate}
depth  <- 10                                          # metres
at_10m <- light_at_depth("clear_noon", "II", depth,
                         wavelength_policy = "trim",
                         surface_source = "direct", surface_angle = 30)
lam_depth <- at_10m$lambda
supported <- lam >= min(lam_depth) & lam <= max(lam_depth)
E_10m  <- at_10m$irradiance
E_sub <- E_sub[supported]
lam <- lam_depth

plot(lam, E_sub, type = "l", col = "goldenrod", lwd = 2,
     xlab = "Wavelength (nm)", ylab = "Irradiance (W/m2/nm)",
     main = "Surface vs 10 m (Jerlov II)")
lines(lam, E_10m, col = "steelblue4", lwd = 2)
legend("topright", c("Just below surface", "10 m"),
       col = c("goldenrod", "steelblue4"), lwd = 2, bty = "n")
```

## 2. Light to capture

Here, quantum catch means the in-water photon irradiance weighted by the
receptor's spectral sensitivity. It remains an area-normalised quantity, not an
absolute per-receptor photon rate. The sensitivity itself is not
the bare pigment template — a real outer segment is partly self-screening, which
broadens the effective curve. `receptor_absorptance()` applies that for a chosen
axial optical density.

```{r capture}
viewer <- "Danio rerio"                               # a tetrachromat
recs   <- unique(subset(species_sensitivities,
                        species == viewer)$receptor)

lam_s  <- seq(300, 750, by = 1)
catch  <- sapply(recs, function(r) {
  row <- subset(species_sensitivities,
                species == viewer & receptor == r)
  tmpl <- govardovskii_template(lam_s, row$lambda_max[1], row$chromophore[1])
  S    <- receptor_absorptance(tmpl, optical_density = 0.4)   # self-screening
  quantum_catch(E_10m, lam, S, input_unit = "W/m2/nm")
})
round(catch)
```

The catches span almost an order of magnitude across the four cones — a direct
readout of how much each receptor's sensitivity overlaps the residual blue-green
field at depth, rather than the broad, even excitation it would receive at the
surface.

## 3. Capture to detection

Now the target. Give it a reddish reflectance and view it against a flat grey
background under the 10 m light field. `inherent_contrast()` returns both an
achromatic (brightness) Weber contrast and a chromatic distance (ΔS, in JNDs)
for the viewer's receptors.

```{r contrast}
grey <- rep(0.30, length(lam))
red  <- 0.20 + 0.25 * exp(-((lam - 625) / 35)^2)

ic <- inherent_contrast(red, grey, E_10m, lam, species = viewer)
ic
```

The default scenario assumes both scalar contrasts decay with distance as
veiling light replaces the object (`C(r) = C0 * exp(-c r)`). This is a teaching
heuristic, not an empirically validated prediction of an animal's actual
detection range. `detectability()` bundles the inherent contrast, the
threshold-distance estimate per channel, and the contrast-vs-distance curve. We pass a real
beam attenuation `c = a + b` rather than the `Kd * kd_to_c` proxy.

```{r detect}
c_beam <- beam_attenuation(absorption = 0.20, scattering = 0.12)   # 1/m
Kd_ref <- jerlov_Kd("II", lambda = 490)

d <- detectability(red, grey, E_10m, lam, Kd = Kd_ref,
                   beam_c = c_beam, species = viewer)
d
```

```{r detect-plot}
plot(d)
```

Under this heuristic the two channels behave very differently. The brightness
(achromatic) difference
stays above threshold for several metres, while the colour signal — only just
above 1 JND at zero distance — washes out almost immediately as veiling light
floods in. Detecting an object by its colour underwater is demanding: it needs
both photons and inherent contrast to spare. Looking **up** (against the brighter
surface) extends the range, because the contrast-attenuation coefficient becomes
`c - Kd`:

```{r looking-up}
detection_range(red, grey, E_10m, lam, Kd = Kd_ref, beam_c = c_beam,
                species = viewer, channel = "achromatic",
                direction = "up")$achromatic
```

For chromatic work, prefer the wavelength-resolved path model. It requires an
explicit beam-attenuation spectrum and veiling radiance rather than treating a
zero-distance JND as a scalar that decays exponentially. In this illustrative
geometry, the grey background is also the asymptotic veiling field:

```{r spectral-path}
Kd_lambda <- jerlov_Kd("II", lambda = lam)
c_lambda  <- 1.5 * Kd_lambda
veil      <- grey * irradiance2radiance(E_10m, geometry = "lambertian")

d_spectral <- detectability(
  red, grey, E_10m, lam,
  Kd = Kd_lambda,
  beam_c = c_lambda,
  species = viewer,
  model = "spectral_path",
  veiling = veil,
  max_distance = 50,
  search_points = 201
)
d_spectral
```

This remains a scenario estimate: the spectral calculation improves the path
physics and receptor calculation, but it does not add target size, acuity,
photon limits, behaviour, or empirical validation for the organism.

## Summary

From a single sunlight spectrum we crossed the surface, propagated to depth,
computed photoreceptor quantum catches (with self-screening), and turned the
object/background spectra into a heuristic threshold-distance scenario — the full
`light → capture → detection` chain, each step a documented luxR function.
