---
title: "From photons to perception"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From photons to perception}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4
)
```

A sensory biologist has spectral reflectance measurements of two substrates and
wants to know whether a fish can discriminate them — and at what depth
discrimination fails as the blue-shifted underwater illuminant narrows the visible
contrast between the two surfaces.

## Units and conversions

We begin with the Naples surface spectrum and demonstrate the unit conversions
central to sensory ecology. The same photon field can be expressed as an energy
flux (W m⁻² nm⁻¹), a photon flux (µmol m⁻² s⁻¹ nm⁻¹), an illuminance (lux), or
a scotopic illuminance.

```{r load}
library(luxR)

x0m   <- from_naples("0m")          # µmol m⁻² s⁻¹ nm⁻¹
x0m_W <- convert_unit(x0m, "W/m2/nm")
x0m_W
```

```{r unit-table}
unit_table <- data.frame(
  Quantity = c(
    "Integrated energy flux in PAR window (W m⁻²)",
    "Photon flux — PAR (µmol m⁻² s⁻¹)",
    "Photopic illuminance (lux)",
    "Scotopic illuminance (scotopic lux)"
  ),
  Value = c(
    round(sum(x0m_W$E[x0m_W$lambda >= 400 &
                      x0m_W$lambda <= 700]) * x0m_W$binwidth, 1),
    round(par_irradiance(x0m_W), 1),
    round(irradiance2lux(x0m_W), 0),
    round(scotopic_lux(x0m_W), 0)
  )
)
knitr::kable(unit_table)
```

Photopic lux weights visible wavelengths by the human eye's sensitivity curve;
scotopic lux uses the dark-adapted (rod) curve. Neither preserves spectral
shape — which matters when asking what a non-human visual system actually detects.

> **Numeric API note:** All functions also accept plain numeric vectors.
> For example, `irradiance2lux(x0m_W$E, x0m_W$lambda)`.

## How the spectrum changes with depth

Using a Jerlov Type II open-water model we propagate the surface spectrum to
five depths and compare radiometric and photometric summaries.

```{r propagate}
x0m_Jerlov <- x0m_W[350, 700]
lam   <- x0m_Jerlov$lambda
Kd_II <- jerlov_Kd("II", lam)
specs <- propagate_spectrum(x0m_Jerlov, Kd_II, from = 0,
                            to = c(0, 5, 10, 20, 50))
```

```{r depth-table}
depth_table <- data.frame(
  depth_m  = c(0, 5, 10, 20, 50),
  PAR_umol = round(sapply(specs, par_irradiance), 1),
  lux      = round(sapply(specs, irradiance2lux), 0)
)
knitr::kable(depth_table,
             col.names = c("Depth (m)",
                           "PAR (µmol m⁻² s⁻¹)",
                           "Photopic illuminance (lux)"))
```

PAR retains spectral shape and photon-flux information relevant to photoreception.
Lux collapses the spectrum into a single number weighted for human daytime vision —
it can mislead when applied to non-human or scotopic scenarios.

## Building a visual model

We use the zebrafish (*Danio rerio*) cone sensitivities retrieved from the package
database via `species_LEF()`. Zebrafish (*Danio rerio*) are tetrachromats with four cone classes (UV, S, M, L), but we restrict the model to the three longer-wavelength cones (S at 415 nm, M at 480 nm, L at 561 nm) because the UV-cone's peak lies at the edge of the modelled illuminant range.

```{r pigments}
sws_df <- species_LEF("Danio rerio", receptor = "S-cone", lambda = lam)
mws_df <- species_LEF("Danio rerio", receptor = "M-cone", lambda = lam)
lws_df <- species_LEF("Danio rerio", receptor = "L-cone", lambda = lam)

plot(sws_df$lambda, sws_df$S, type = "l", col = "blue",
     xlab = "Wavelength (nm)", ylab = "Normalised absorbance",
     main = "Zebrafish (Danio rerio) cone sensitivities",
     ylim = c(0, 1))
lines(mws_df$lambda, mws_df$S, col = "green3")
lines(lws_df$lambda, lws_df$S, col = "red3")
legend("topright",
       legend = c("S-cone (415 nm)", "M-cone (480 nm)", "L-cone (561 nm)"),
       col = c("blue", "green3", "red3"), lty = 1)
```

`quantum_catch()` integrates photon irradiance weighted by dimensionless
sensitivity over wavelength. Its result is in photons m⁻² s⁻¹ at the
measurement plane; it is not an absolute photons-per-second rate for one
receptor. It takes a numeric irradiance vector with an explicit unit, so we
extract `$E` from each depth spectrum and declare its energy unit.

```{r quantum-catch}
qc_list <- lapply(specs, function(s) {
  data.frame(
    S = quantum_catch(s$E, s$lambda, sws_df,
                      input_unit = "W/m2/nm", binwidth = s$binwidth),
    M = quantum_catch(s$E, s$lambda, mws_df,
                      input_unit = "W/m2/nm", binwidth = s$binwidth),
    L = quantum_catch(s$E, s$lambda, lws_df,
                      input_unit = "W/m2/nm", binwidth = s$binwidth)
  )
})
qc_df         <- do.call(rbind, qc_list)
qc_df$depth_m <- c(0, 5, 10, 20, 50)
qc_df$M_L     <- round(qc_df$M / qc_df$L, 3)

knitr::kable(qc_df[, c("depth_m", "S", "M", "L", "M_L")],
             digits = 2,
             col.names = c("Depth (m)", "S-cone catch",
                           "M-cone catch", "L-cone catch", "M/L ratio"))
```

As depth increases the M/L ratio rises: the blue-shifted deep illuminant peaks
near 500 nm, which aligns with the M-cone (480 nm) better than the L-cone
(561 nm), shifting the chromatic balance available to the fish.

## Reflectance to radiance

We define two synthetic substrates: a spectrally flat grey (reflectance = 0.3
everywhere) and a brownish-red substrate with a Gaussian peak near 625 nm.

```{r reflectances}
grey_R <- lux_spectrum(rep(0.3, length(lam)), lam,
                       quantity = "reflectance", unit = "dimensionless")
red_E  <- 0.2 + 0.25 * exp(-((lam - 625) / 35)^2)
red_R  <- lux_spectrum(red_E, lam,
                       quantity = "reflectance", unit = "dimensionless")

plot(grey_R$lambda, grey_R$E, type = "l", col = "grey50",
     ylim = c(0, 0.6),
     xlab = "Wavelength (nm)", ylab = "Reflectance",
     main = "Substrate reflectance spectra")
lines(red_R$lambda, red_R$E, col = "firebrick")
legend("topright", legend = c("Grey (0.3)", "Red (625 nm peak)"),
       col = c("grey50", "firebrick"), lty = 1)
```

`reflectance_to_radiance()` multiplies each reflectance by the depth-dependent
illuminant to give the stimulus reaching the observer's eye.

```{r radiance}
rad_grey_0m  <- reflectance_to_radiance(grey_R, specs[["0"]])
rad_red_0m   <- reflectance_to_radiance(red_R,  specs[["0"]])
rad_grey_20m <- reflectance_to_radiance(grey_R, specs[["20"]])
rad_red_20m  <- reflectance_to_radiance(red_R,  specs[["20"]])

ylim_rad <- range(c(rad_red_0m$E, rad_red_20m$E,
                    rad_grey_0m$E, rad_grey_20m$E))
plot(rad_red_0m$lambda, rad_red_0m$E, type = "l", col = "firebrick",
     ylim = ylim_rad,
     xlab = "Wavelength (nm)",
     ylab = expression("Radiance (W m"^{-2}~"nm"^{-1}*")"),
     main = "Substrate radiances at surface vs 20 m")
lines(rad_grey_0m$lambda,  rad_grey_0m$E,  col = "grey50")
lines(rad_red_20m$lambda,  rad_red_20m$E,  col = "firebrick", lty = 2)
lines(rad_grey_20m$lambda, rad_grey_20m$E, col = "grey50",    lty = 2)
legend("topright",
       legend = c("Red (0 m)", "Grey (0 m)",
                  "Red (20 m)", "Grey (20 m)"),
       col = c("firebrick", "grey50", "firebrick", "grey50"),
       lty = c(1, 1, 2, 2))
```

At 20 m the two radiance curves converge: the blue-shifted illuminant strongly
attenuates the red substrate's 625 nm peak, leaving it nearly indistinguishable
from the flat grey.

## Colour discrimination

`colour_jnd()` implements the Vorobyev–Osorio receptor-noise-limited model. A
JND ≥ 1 means the pair is reliably discriminable; JND < 1 means it is not. We use
zebrafish (*Danio rerio*) S-, M- and L-cone sensitivities with a Weber fraction of
0.05, comparing the grey and red substrates. When `receptor` is omitted,
`colour_jnd()` uses the species' cited default chromatic channel from
`species_channels`; it never promotes rods, irradiance detectors, or incomplete
receptor sets to colour channels.

`colour_jnd()` takes numeric radiance vectors (W m⁻² nm⁻¹), so we extract `$E`
and `$lambda` from the `lux_spectrum` objects.

```{r colour-jnd}
depths     <- c(0, 5, 10, 20, 50)
depth_keys <- as.character(depths)

jnd_vals <- sapply(depth_keys, function(d) {
  r1 <- reflectance_to_radiance(grey_R, specs[[d]])
  r2 <- reflectance_to_radiance(red_R,  specs[[d]])
  colour_jnd(
    stim1    = r1$E,
    stim2    = r2$E,
    lambda   = r1$lambda,
    species  = "Danio rerio",
    receptor = c("S-cone", "M-cone", "L-cone"),
    noise    = 0.05,
    binwidth = r1$binwidth
  )
})

jnd_table <- data.frame(depth_m = depths, JND = round(jnd_vals, 2))
knitr::kable(jnd_table,
             col.names = c("Depth (m)", "ΔS (JND)"))
```

```{r jnd-plot}
plot(jnd_table$depth_m, jnd_table$JND, type = "b", pch = 16,
     xlab = "Depth (m)",
     ylab = expression(Delta * S ~ (JND)),
     main = "Colour discrimination vs depth")
abline(h = 1, lty = 2, col = "red")
text(35, 1.3, "discrimination\nthreshold", col = "red", cex = 0.85)
```

```{r threshold}
fail_idx <- which(jnd_table$JND < 1)
if (length(fail_idx) > 0) {
  cat(sprintf(
    "Discrimination fails (JND < 1) at %d m depth.\n",
    jnd_table$depth_m[fail_idx[1]]
  ))
} else {
  cat("JND remains above 1 at all modelled depths.\n")
}
```

The model predicts that the zebrafish can discriminate the grey and red substrates
in well-lit shallow water (JND ≈ 2–3 near the surface) but loses the ability
by 20 m, where the blue-shifted Jerlov II illuminant has attenuated the red
substrate's 625 nm peak sufficiently to render it indistinguishable from the flat
grey (JND < 1).

## Validation against pavo

luxR's `colour_jnd()` is the Vorobyev–Osorio receptor-noise-limited model, the
same model implemented by the widely used [pavo](https://CRAN.R-project.org/package=pavo)
package (`vismodel()` + `coldist()`). Given identical receptor sensitivities,
stimuli, and noise, the two must agree. They do — to machine precision — once the
conventions are matched:

* feed pavo luxR's own Govardovskii templates as the visual system;
* luxR integrates **photon** catch (∝ `E × lambda`); reproduce that in pavo with an
  illuminant proportional to wavelength;
* luxR's per-receptor noise is the Weber fraction directly; reproduce that with
  equal cone ratios (`n = 1`), so `e_i = weber`;
* use raw quantum catches: `vismodel(qcatch = "Qi", vonkries = FALSE, relative = FALSE)`.

```{r pavo-validation, eval = requireNamespace("pavo", quietly = TRUE)}
library(pavo)

lam <- 300:750
s1  <- approx(solar_spectra$clear_noon$wavelength,
              solar_spectra$clear_noon$irradiance, lam, rule = 2)$y
s2  <- approx(solar_spectra$underwater_10m$wavelength,
              solar_spectra$underwater_10m$irradiance, lam, rule = 2)$y
w   <- 0.05

pavo_dS <- function(species) {
  members <- species_channels[
    species_channels$species == species &
      species_channels$channel_role == "chromatic" &
      species_channels$is_default,
  ]
  species_rows <- species_sensitivities[
    species_sensitivities$species == species,
  ]
  recs <- species_rows[match(members$receptor, species_rows$receptor), ]
  S <- sapply(seq_len(nrow(recs)), function(i)
    govardovskii_template(lam, recs$lambda_max[i], recs$chromophore[i])$S)
  colnames(S) <- make.unique(recs$receptor)
  vm <- vismodel(as.rspec(data.frame(wl = lam, s1 = s1, s2 = s2), lim = range(lam)),
                 visual = as.rspec(data.frame(wl = lam, S), lim = range(lam)),
                 illum = lam, qcatch = "Qi", vonkries = FALSE,
                 relative = FALSE, achromatic = "none")
  coldist(vm, noise = "neural", n = rep(1, nrow(recs)),
          weber = w, weber.ref = 1, achromatic = FALSE)$dS[1]
}

species <- c("Homo sapiens", "Apis mellifera", "Danio rerio")
cmp <- data.frame(
  species = species,
  luxR    = sapply(species, function(sp)
              colour_jnd(s1, s2, lambda = lam, species = sp, noise = w)),
  pavo    = sapply(species, function(sp) suppressMessages(pavo_dS(sp)))
)
cmp$abs_diff <- abs(cmp$luxR - cmp$pavo)
knitr::kable(cmp, digits = c(0, 6, 6, 14),
             col.names = c("Species", "luxR ΔS", "pavo ΔS", "|difference|"))
```

The agreement is a useful regression guard as well as a check on correctness:
`tests/testthat/test-pavo-validation.R` asserts it automatically wherever pavo is
installed.

## Interoperating with pavo

Beyond validation, luxR and pavo *compose*: luxR propagates the light field
through the water, and pavo analyses the resulting colours. `as_rspec()` converts
luxR spectra into pavo's `rspec` class, after which the whole pavo toolkit
(`vismodel()`, `coldist()`, `colspace()`, plotting) is available; an
`as_lux_spectrum()` method converts back.

The pipeline below propagates daylight to 15 m in Jerlov II water with luxR, then
uses that **in-water light field as the illuminant** for a pavo visual model of
two substrates — a calculation neither package does alone.

```{r pavo-interop, eval = requireNamespace("pavo", quietly = TRUE)}
library(pavo)

# luxR: the downwelling light field at 15 m (this becomes pavo's illuminant)
Ld  <- light_at_depth("clear_noon", "II", depth = 15,
                      wavelength_policy = "trim")

# Two object reflectances, bridged into pavo via as_rspec() (interpolated to 1 nm)
lam   <- Ld$lambda
gauss <- function(centre) 0.2 + 0.25 * exp(-((lam - centre) / 35)^2)
refl  <- as_rspec(list(
  red_substrate  = as_lux_spectrum(
    gauss(625), lambda = lam,
    quantity = "reflectance", unit = "dimensionless"
  ),
  grey_substrate = as_lux_spectrum(
    rep(0.3, length(lam)), lambda = lam,
    quantity = "reflectance", unit = "dimensionless"
  )
), lim = c(300, 700))   # match pavo's built-in honeybee visual system

# Put luxR's light field on the rspec's (1 nm) grid so it can serve as the illuminant
illum <- approx(Ld$lambda, Ld$irradiance, refl$wl, rule = 2)$y

# pavo: model a honeybee viewing those substrates UNDER luxR's underwater light
vm <- vismodel(refl, visual = "apis", illum = illum,
               qcatch = "Qi", relative = FALSE)
coldist(vm, n = c(1, 1, 1), achromatic = FALSE)[, c("patch1", "patch2", "dS")]
```

luxR owns the water; pavo owns the colour space, and `as_rspec()` is the seam.

## Interoperating with colourvision

luxR also bridges into
[colourvision](https://CRAN.R-project.org/package=colourvision), which
implements receptor-noise-limited and colour-space models. As with pavo, luxR
propagates the in-water light field and colourvision analyses the resulting
colours. colourvision has no dedicated spectral class: each model argument is a
plain data frame whose first column is wavelength. `as_colourvision()` produces
that shape, and `species_sensitivity_matrix()` builds the receptor matrix `C`
from luxR's own Govardovskii templates.

```{r cv-interop, eval = requireNamespace("colourvision", quietly = TRUE)}
library(colourvision)

lam <- 300:700

# luxR: the downwelling light field at 15 m in Jerlov II water becomes
# colourvision's illuminant I.
Ld <- light_at_depth("clear_noon", "II", depth = 15, wavelength_policy = "trim")
I <- as_colourvision(
  as_lux_spectrum(
    stats::approx(Ld$lambda, Ld$irradiance, lam, rule = 2)$y,
    lambda = lam, quantity = "irradiance", unit = "W/m2/nm"
  ),
  name = "I")

# Two object reflectances as R1 / R2, flat background Rb.
R1 <- data.frame(wl = lam, R = 0.2 + 0.25 * exp(-((lam - 550) / 30)^2))
R2 <- data.frame(wl = lam, R = 0.2 + 0.25 * exp(-((lam - 600) / 30)^2))
Rb <- data.frame(wl = lam, R = rep(0.1, length(lam)))

# luxR's own receptors as colourvision's C (photon-weighted).
C <- species_sensitivity_matrix("Apis mellifera", lambda = lam)

RNLmodel(model = "log", R1 = R1, R2 = R2, Rb = Rb, I = I, C = C,
         noise = TRUE, e = rep(0.05, ncol(C) - 1),
         interpolate = FALSE, nm = lam)[["deltaS"]]
```

luxR owns the water; colourvision owns the colour space, and `as_colourvision()`
plus `species_sensitivity_matrix()` are the seam.

### Validation against colourvision

luxR's `colour_jnd()` and colourvision's `RNLmodel(model = "log")` implement the
same Vorobyev-Osorio model. Given matched sensitivities and noise they agree to
numerical precision: the photon-vs-energy integration difference is absorbed by
`species_sensitivity_matrix(weight = "photon")` (which emits `S`&nbsp;`*`&nbsp;`lambda`),
and colourvision's background von Kries factor cancels in the pairwise log
difference. `tests/testthat/test-colourvision-validation.R` asserts this
automatically wherever colourvision is installed.
